Reproducible Research. Peer Assessment 1
========================================================

R Markdown document for the Peer Assessment 1  
Part of the course "Reproducible Research"

## Introduction
This assignment makes use of data from a personal activity monitoring  device. This device collects data at 5 minute intervals through out the day.

The data for this assignment can be downloaded from the [course web site] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  


## Loading and preprocessing the data

**Unload & unzip**
```{r echo = TRUE}
setInternet2(TRUE) # avoid error, source: R documentation (?download.file()
setwd("~/R")
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file (fileURL, destfile = "~/R/data.zip")
unzip("data.zip")
```

**Read & explore**
```{r reading, echo = TRUE}
data.act <- read.csv("activity.csv", nrows = 17569) # import the file w/o changes
````
These commands will be used repeatedly throughout the work
```{r exploring, echo = TRUE}
head(data.act)
tail(data.act)
summary(data.act)
str(data.act)
````

Load the libraries that I use for the job
```{r libraries, echo=TRUE}
library(ggplot2)
library(knitr)
library(htmltools)
library(rmarkdown)
```

**Creates a new dataframe**  
accumulated steps per date
```{r echo = TRUE}
steps.date <- aggregate(steps ~ date, data.act, sum)
```



## What is mean total number of steps taken per day?

**Make a histogram of the total number of steps taken each day**  
clarification: x = total number of steps per day, y = frequency)
```{r firstplot, echo = TRUE}
ggplot(steps.date, aes(x=steps)) + geom_histogram(binwidth=500) + labs(title = "Total number \nof steps taken each day")
```

**Calculate and report the mean and median total number of steps taken per day**
```{r echo = TRUE}
paste ("The mean is", round(mean(steps.date$steps),2), "steps per day")
paste ("The median is", median(steps.date$steps), "steps per day")
```



## What is the average daily activity pattern?  

**Make a time series plot of 5-minute interval and the average of steps taken, averaged across all days**
```{r secondplot, echo = TRUE}
steps.interval <- aggregate(steps ~ interval, data.act, mean)
ggplot(steps.interval, aes(x=interval, y=steps)) + geom_line() + ggtitle("Average daily activity") + ylab("Average number steps")
```

**Which 5-minute interval, on average across all the days in the dataset contains the maximum number of steps?**
```{r echo = TRUE}
paste("The highest number of steps corresponds to the interval", steps.interval[which.max(steps.interval$steps),][,1], "with", round(steps.interval[which.max(steps.interval$steps),][,2],2), "steps")
```



## Imputing missing values

**Calculate and report the total number of missing values**
```{r echo =TRUE}
paste("There are", sum(is.na(data.act$steps)), "rows with NA steps")
paste("There is", sum(is.na(data.act$date)), "rows with NA dates")
paste("There is", sum(is.na(data.act$interval)), "rows with NA intervals")
```

**Devise a strategy for filling in all of the missing values in the dataset**  
strategy: replace NA values with the mean value of the same interval  
  
join two datasets by the variable "interval"
```{r echo = TRUE}
fill.data <- merge(data.act, steps.interval, by = "interval", sort = FALSE)
```

rename column names
```{r echo =TRUE}
colnames(fill.data) <- c("interval","steps","date","mean.steps.interval")
```

**Create a new dataset that is equal to the original dataset but with the NA values filled in**
```{r echo = TRUE}
fill.data$steps[is.na(fill.data$steps)] <- fill.data$mean.steps.interval[is.na(fill.data$steps)]
```

clean dataframe, erase last column (optional)
```{r echo =TRUE}
fill.data$mean.steps.interval <- NULL
```

**Make a histogram of the total number of steps taken each day**
```{r thirdplot, echo = TRUE}
fill.steps.date <- aggregate(steps ~ date, fill.data, sum)
ggplot(fill.steps.date, aes(x=steps)) + geom_histogram(binwidth=500) + ggtitle("Total number of days according to steps") + ylab("frecuency")
```

**Calculate and report the mean and median total number of steps taken per day**
```{r echo = TRUE}
paste("The new mean is", round(mean(fill.steps.date$steps),2), "steps by day")
paste("The new median is", round(median(fill.steps.date$steps),2), "steps by day")
```

**Do these values differ from the estimates from the first part of the assignment?**  

```{r echo=TRUE}
paste("The difference between the old and the new mean is equal to", round(mean(steps.date$steps) - mean(fill.steps.date$steps),2))
```


```{r echo = TRUE}
paste("The difference between the old and the new median is equal to", round(median(steps.date$steps) - median(fill.steps.date$steps),2))
``` 

**What is the impact of imputing missing data on the estimates of the total daily number of steps?**  
We observe that the mean no changes and the new median is larger than the last, approaching the mean



## Are there differences in activity patterns between weekdays and weekends?

**Create dataset with the new factor**
```{r echo=TRUE}
day.data <- fill.data # duplicate dataset (optional)
day.data$days <- as.factor(weekdays(as.Date(day.data$date))) # creates new factor
levels(day.data$days) <- list(weekday = c("lunes", "martes", "miércoles","jueves", "viernes"), weekend = c("sábado", "domingo")) # fix levels of the new factor
``` 
*The names of the list depends de local language,*
*alternatively, it can be fixed to english with*
*Sys.setlocale("LC_TIME", "English") and the list is in english words.*

**Make a panel plot of the 5-minute interval (x-axis) and the average number of steps taken**

```{r lastplot, echo= TRUE}  
week.data <- aggregate(steps ~ interval+days, day.data, mean)
p <- ggplot(week.data, aes(x=interval, y=steps)) + geom_line() + ylab("average steps") + ggtitle("daily activity") 
p + facet_grid( .~days)
```

We can see that there are different patterns of behavior, associated, probably, with work schedules.